---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
date: "2025-09-25"
author:
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: 0000-0001-9012-3060
    affiliation: Maynooth University
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: 0000-0002-0875-3563
    affiliation: Maynooth University
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: 0000-0001-6159-6707
    affiliation: National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: 0000-0002-7387-3107
    affiliation: Maynooth University
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging
published: "This manuscript is a preprint and is currently under review at *Alzheimer's & Dementia: Diagnosis, Assessment & Disease Monitoring*"
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
editor:
  markdown:
    wrap: 80
keep-tex: true
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(tinytable)

knitr::opts_chunk$set(echo = FALSE, include = FALSE,
warning = FALSE, message = FALSE)

knitr::opts_chunk$set(
  fig.align = "center", fig.retina = 3,
  fig.width = 6, fig.height = (6 * 0.618),
  out.width = "100%", collapse = TRUE
)

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 10),
      plot.caption = ggtext::element_markdown(hjust = 1),
      axis.title = element_text(size = 10, face = "bold"),
      axis.text = element_text(size = 9),
      strip.text = element_text(size = 9, face = "bold"),
      panel.grid = element_blank()
    )
  )
```

::: {.content-visible when-format="html"}
# Introduction {.hide}
:::

Dementia is characterized by progressive decline of cognitive function, leading
to memory loss and difficulties with daily living [@prince2013;
@sanz-blasco2022]. The global burden of dementia is substantial, with cases
projected to rise to 152.8 million by 2050 [@nichols2022]. Identifying and
addressing modifiable risk factors is crucial to mitigate this projected rise in
prevalence. Two particularly vulnerable groups are older adults and those with
mild cognitive impairment (MCI), of whom up to 80% progress to dementia within
six years [@cooper2015; @shigemizu2020 @tschanz2006].

The Lancet Commission on Dementia identified fourteen causal risk factors
including smoking, hypertension, and diabetes, which account for approximately
45% of dementia cases worldwide [@livingston2024]. Prodromal markers, on the
other hand, reflect early disease processes and may signal limited opportunities
for prevention [@teipel2025]. One such prodromal marker is apathy, or the
significant loss of motivation, as distinct from depression and cognitive
impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
dementia subgroups [@donovan2014; @fresnais2023, @salem2023; @vandalen2018].
Apathy is a significant predictor of the transition from MCI to dementia
[@ruthirakuhan2019; @salem2023, @vandalen2018]. While apathy has been
well-established as a prodromal marker and risk factor for dementia progression,
emerging evidence suggests that procrastination may also relate to early
cognitive decline.

Though superficially similar, apathy and procrastination are behaviorally
distinct. Apathy reflects reduced internal drive and emotional engagement,
impairing initiation of action [@fresnais2023], whereas procrastination reflects
intact intention but delayed execution [@steel2007]. While both reflect
impairments in executive functioning, procrastination may additionally capture
unique aspects of emotional and motivational dysregulation . For instance,
person A may feel no motivation to attend an exercise class and therefore never
books it (apathy). Person B, however, books the class and intends to go, but
postpones at the last minute (procrastination). Both apathy and procrastination
have been linked to dysfunction in prefrontal regions implicated in dementia
[@friden2020; @zhang2010; @joseph2021; @fahed2021].

These parallels raise the possibility that procrastination could serve as an
early marker of cognitive impairment or even a modifiable risk factor. Chronic
procrastination may exacerbate decline by limiting engagement in cognitively
stimulating activities such as physical exercise, problem-solving, and
goal-setting, which build cognitive resilience and reduce dementia risk
[@kelly2021; @mohammadibytamar2020; @chowdhary2022; @livingston2024]. Reduced
engagement may contribute to a cycle of cognitive disengagement, accelerating
decline.

Importantly, procrastination is responsive to intervention through
cognitive-behavioral strategies and self-regulation training [@rozental2014;
@vaneerde2018]. Crucially, procrastination may identify individuals with
motivational or executive dysfunction who do not yet meet the long-term symptom
threshold for apathy, offering a potentially valuable and earlier marker (and
possible target) for neurodegenerative risk.

Since age remains the strongest predictor of MCI and dementia, it is important
to consider how procrastination operates across the lifespan (i.e. interaction
with age). Many established modifiable risk factors such as hypertension,
hearing loss, smoking, and social isolation, exert age-dependent effects, with
certain factors carrying more weight at midlife than in late life
[@livingston2020; @livingston2024]. Understanding how procrastination interacts
with age may help clarify its role in the aetiology of cognitive decline and
identify windows of opportunity for targeted intervention.

To our knowledge, no studies have directly examined procrastination as a
predictor of cognitive decline or dementia progression. Accordingly, the present
study aimed to a) assess differences in procrastination levels across three
groups: individuals with dementia, individuals with MCI, and individuals with
neither dementia nor MCI, and b) test whether higher procrastination scores
predict transition from normative cognition function to dementia or MCI to
dementia.

# Method

## Data and study population

Analyses were conducted using a secondary data source; a multi-wave prospective
cohort study called the Health and Retirement Study (HRS; [@fisher2018], which
tracks the health, economic, and social well-being of over $18,000$ American
adults primarily aged $50+$. The HRS is managed by the Institute for Social
Research at the University of Michigan, with data collected every two years.
Initial data collection of a participant is conducted through a face-to-face
interview, with follow-up biennial interviews conducted either by phone or
face-to-face. The average retention rate ranges from $68.8\%$ to $92.3\%$
[@HRS2017]. At the time of writing, fifteen years of HRS data are currently
archived.

We focused on four waves of HRS data from 2016 to 2022. Specifically, our study
sample consisted of respondents who participated in an experimental module
assessing procrastination during the 2020 wave. These experimental modules,
administered at the end of the core HRS interview, consist of concise
questionnaires designed to explore new topics or supplement existing core survey
data [@juster1995]. Each respondent receives only one experimental module, with
sample sizes for each module constituting approximately 10% of the core sample.
As a result, while the core HRS sample includes approximately $18,000$
respondents, our initial sample of interest consisted of $1,368$ respondents. We
excluded respondents with missing cognitive assessment data for any wave
$(n = 419)$, those under 60 years of age, (as cognitive symptoms typically occur
around this age $n = 398$), and those with complete missing values across the
procrastination measure $(n = 2)$. This resulted in a final analytic sample of
$549$ respondents.

## Measures

### Outcome: Cognitive Function and Cognitive Category

Cognitive function in the HRS is assessed using a series of tests adapted from
the Telephone Interview for Cognitive Status (TICS; [@brandt1988; @fong2009].
These tests include an immediate and delayed $10$-noun free recall test (to
assess episodic memory), a serial seven subtraction test (to assess working
memory), and a backward count from $20$ test (to assess mental processing).
Based on these assessments, @crimmins2011 developed both a $27$-point cognitive
scale and validated cut-off points to assess and classify cognitive status.
Using these points, respondents who scored $12 - 27$ were classified as having
normal cognition, $7 - 11$ as having MCI, and $0 - 6$ as having dementia.

### Predictor: Procrastination

Procrastination was measured using the Pure Procrastination Scale [@steel2010],
a psychometrically validated scale for evaluating procrastination when
conceptualized as a dysfunctional delay. The PPS consists of 12 items rated on a
Likert scale ranging from $1$ (strongly disagree) to $5$ (strongly agree). In
responding to the scale items, participants were asked to reflect on their
general behavioural tendencies, with no specific time-frame provided. Total
procrastination scores range from $12$ to $60$, with higher scores indicating
greater tendency to procrastinate. In the HRS, the PPS was administered only in
wave 3 (2020) of the respective waves. As such, we use the wave 3 measure as
both a retrospective and prospective proxy for procrastination scores across all
waves in the analysis. This approach assumes relative temporal stability in
procrastination over the study period. In this sample, the PSS had a Cronbach’s
$\alpha$ score of 0.92, indicating high internal consistency. In responding to
the scale items, participants were asked to reflect on their general behavioural
tendencies, with no specific time-frame provided. An example of a question from
the scale includes \`\`I delay making decisions until it’s too late”.

### Covariate: Apathy

While no direct measure of apathy exists within the HRS, we utilised two
questions from the eight-item version of the Center for Epidemiological Studies
Depression (CES-D8) scale [@briggs2018] as proxies for apathy: "You felt that
everything you did was an effort" and "You could not get going". Both items
capture core features of apathy (behavioural and motivational disengagement)
and, while not a comprehensive measure of apathy, provide a valid and pragmatic
approximation for modelling purposes. Each item was measured on a binary
“yes/no” scale with total scores ranging from \$0\$ to \$2\$.

### Confounders

To account for potential confounding, we controlled for demographic variables
with established associations with both cognitive function and procrastination
[@vanderweele2019]. These included measures of age, sex, and educational
attainment [@abner2012; @freedman2024; @yu2013]. Educational attainment was
classified into three categories: no formal education, GED (General Educational
Development)/high school diploma, and college/further education.

\clearpage

## Data Analysis

All data analysis was carried out in R [@rcoreteam2025]. To model transitions in
cognitive states over time, we employed a first order discrete-time Markov
model, a class of stochastic processes that satisfy the Markov property
[@zhang2010], which can be formally expressed as:

$$P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)$$

This property asserts that the probability of transitioning from state $X_t = i$
to a future state $X\_{t+1} = j$ depends only on the current state $X_t$, and
not on the full history of preceding states. Here, there are three potential
states: normative cognitive function, mild cognitive impairment, and dementia.
We consider dementia to be an absorbing state, i.e. once an individual reaches
this state, they cannot return to the other states in a future time point.

Unlike continuous-time models, discrete-time Markov models are not readily
supported by a dedicated R package for deriving transition probabilities.
Therefore, we implemented the model manually using multinomial logistic
regression from the *nnet* package [@venables2002]. This approach estimates the
log-odds of transitioning to each non-reference state as a linear function of
covariates, relative to a chosen reference category. For a system with $K$
cognitive states (with state $K$ as the reference), the model takes the form:

$$log \frac{P(Y = j \; \vert \; x)}{P(Y = K \vert x)} = \beta_0 + \beta_j^\top x \qquad \text{for } j = 1, \dots K-1$$

From these equations, the predicted transition probabilities for the
non-reference states $j = 1, \dots, K - 1$ are derived as:

$$P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^\top x}}{1 + \sum^{K - 1}_{k = 1} e^{\beta_{0k} + \beta_k^ \top x}}$$

and for the reference state $K$ as:

$$ P(Y = K \; \vert \; x) = \frac{1}{1 + \sum^{K - 1}_{k = 1} e^{\beta_{0k} + \beta_j^ \top x }} $$

To assess model fit and guide model selection for covariates in the $\beta_j$
vectors, we conducted likelihood-ratio tests comparing full and reduced models,
whose test statistic is defined as:

$$ D = 2 \times (\ell_{\text{reduced}} - \ell_{\text{full}}) $$

where $\ell_{\text{full}}$ and $\ell_{\text{reduced}}$ denote the
log-likelihoods of the respective models. Under the null hypothesis that the
full model does not represent an improved fit when compared to the reduced
model, $D$ has an asymptotic $\chi^2$ distribution with degrees of freedom
corresponding to the difference between the numbers of parameters estimated by
the two models.

# Results

Our final analytic sample comprised $549$ respondents with the following age
distribution: 60 - 70 $(n = 186)$, 71 - 80 $(n = 203)$, 81 - 90 $(n = 142)$, and
90+ $(n = 18)$. Descriptive statistics for the full sample, as well as data
stratified by cognitive status (normative cognitive function, MCI, and
dementia), are presented in @tbl-descriptives. Both @fig-river and supplementary
figure S1 capture the unconditional transitions and transition probabilities
(respectively) between wave one and two (first transition), wave two and three
(second transition), and wave three and four (third transition), yielding a
total of $1,647$ observed transitions over time.

```{r}
#| label: tbl-descriptives
#| fig-width: 12
#| tbl-cap: "Baseline descriptives for the study sample and stratified by cognitive status category."
#| tbl-cap-location: top
#| tbl-pos: H
#| include: true

# Table information
tbl_descriptives <- tibble::tribble(
  ~""                 , ~Full              , ~NC                , ~MCI              , ~Dementia        ,
  "Age (years)"       , "69.70 ± 7.58"    , "69.70 ± 7.55"    , "69.60 ± 7.55"   , "70.00 ± 9.89"  ,
  "Male"              , "38.43% (n = 211)" , "39.82% (n =180)"  , "32.56% (n = 28)" , "27.27% (n = 3)" ,
  "Female"            , "61.57% (n = 338)" , "60.18% (n = 272)" , "67.44% (n = 58)" , "72.73% (n = 8)" ,
  "No degree"         , "16.24% (n = 98)"  , "10.56% (n = 47)"  , "39.53% (n = 34)" , "63.54% (n = 7)" ,
  "GED"               , "51.58% (n = 279)" , "51.91% (n = 231)" , "52.33% (n = 45)" , "27.27% (n = 3)" ,
  "Further education" , "32.29% (n = 175)" , "37.53% (n = 167)" , "8.14% (n = 7)"   , "9.09% (n = 1)"  ,
  "Apathy"            , "0.37 ± 0.63"     , "0.31 ± 0.59"     , "0.65 ± 0.76"    , "0.36 ± 0.51"   ,
  "Procrastination"   , "28.60 ± 12.00"   , "27.70 ± 11.30"   , "32.10 ± 13.80"  , "39.00 ± 16.00"
)

# Captions and footnotes
caption <- glue::glue(
  "Note: Descriptives for continuous and categorical variables are represented
  using means ± standard deviations and percentages and frequencies respectively.
  NC = Normative cognitive function; MCI = Mild cognitive impairment;
  GED = General Educational Development.")

footnote <- glue::glue(
  "Procrastination scores were collected in 2020 (Wave 3) and are presented
  here for descriptive comparison, although they were not measured at
  baseline.")

# Outputting table
tbl_descriptives |>
  setNames(c(" ", "Full sample (n = 549)", "NC (n = 452)", "MCI (n = 86)", "Dementia (n = 11)")) |>
  tt(notes = list(
    caption,
    a = list(i = 10, j = 1, text = footnote))
  ) |>
   group_tt(i = list("Sex" = 2, "Education" = 4)) |>
   style_tt(j = 2:5, align = "c") |>
   style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
   format_tt(escape = TRUE) |>
   format_tt("notes", markdown = TRUE) |>
   style_tt(
     bootstrap_class = "table table-sm"
     )
```

```{r}
#| label: fig-river
#| fig-width: 15
#| fig-height: 10
#| fig-cap: "Frequency of cognitive status transitions between HRS waves."
#| fig-pos: H
#| include: true

fig_1 <- readRDS(here::here("files/figure_1.RDS"))

fig_1
```

## Cross sectional differences

Initially, a Kruskal-Wallis test was conducted to examine differences in
procrastination scores (measured in 2020) across three cognitive status groups
after Levene's test indicated violation of homogeneity of variance
$(p = 0.039)$. The analysis revealed a statistically significant effect of
cognitive status, $(\chi^2(2) = 17.54, p < .001)$, indicating that
procrastination levels differed significantly between at least two of the
groups. Post-hoc analysis using a pairwise Wilcoxon rank-sum test with a
Benjamini-Hochberg correction showed that participants with normal cognition
$(M = 27.7, SD = 11.7)$ reported significantly lower levels of procrastination
than those with both MCI $(M = 32.1; SD = 11.3; p = 0.004)$ and dementia
$(M = 36.2; SD = 14.8; p = 0.005)$. No significance difference was found between
those with MCI and dementia $(p = 0.334)$. @fig-cross-sectional displays the
distribution of procrastination scores across groups, with both boxplots and
dotplots showing median values and individual data points. Significance bars
indicate the pairwise differences described above.

```{r}
#| label: fig-cross-sectional
#| fig-width: 10
#| fig-cap: "Group differences in 2020 procrastination scores according to cognitive status."
#| include: true

fig_2 <- readRDS(here::here("files/figure_2.RDS"))

fig_2
```

## Markov analysis

Results from the discrete-time Markov analysis, showed that all covariates
significantly influenced the likelihood of transitioning between cognitive
states (see @fig-markov). Notably, procrastination interacted significantly with
*age* to affect two key transitions: increasing the likelihood of transitioning
from normative cognitive function to MCI $(\text{OR} = 1.001; p < 0.001)$ and
decreasing the likelihood of reverting from MCI to normative cognitive function
$(\text{OR} = 0.999; p < 0.001)$. *Women* were significantly less likely than
men to transition from both normative cognitive function to dementia
$(\text{OR} = 0.068; p < 0.001)$ and from MCI to dementia
$(\text{OR} = 0.70; p < 0.001$).

For *education attainment*, individuals with a GED were less likely to
transition from normative cognitive function to either MCI
$(\text{OR} = 0.49; p < 0.001)$ or dementia $(\text{OR} = 0.29; p < 0.001)$ and
from MCI to dementia $(\text{OR} = 0.64; p < 0.001)$. They were also more likely
to back transition from MCI to normative cognitive function
$(\text{OR} = 2.18; p < 0.001)$. Those with a college level education or higher
demonstrated a significantly reduced likelihood of transitioning from normative
cognitive function to either MCI $(\text{OR} = 0.32; p < 0.001)$ or dementia
$(\text{OR} = 0.07; p < 0.001)$ and from MCI to dementia
$(\text{OR} = 0.22; p < 0.001)$. They were also more likely to back transition
from MCI to normative cognitive function $(\text{OR} = 3.13; p < 0.001)$.

Finally, higher levels of *apathy* were associated with an increased likelihood
of transitioning from normative cognitive function to both MCI
$(\text{OR} = 1.35; p = 0.005)$ and dementia $(\text{OR} = 1.36; p<0.001)$ and a
decreased likelihood of transitioning from MCI back to normative cognitive
function $(\text{OR} = 0.74; p = 0.003)$.

```{r}
#| label: fig-markov
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Odds ratios (95% Confidence Intervals) of transitioning from one cognitive state to another calculated using generalised logit models. *Note*. Odds ratios significantly different from 1 at a 5% significance rate are presented in blue (positive) or orange (negative),  otherwise they are presented in gray. NC = Normative cognitive function;  MCI = Mild cognitive impairment."
#| include: true

fig_3 <- readRDS(here::here("files/figure_3.RDS"))

fig_3
```

@fig-age-preds presents the predicted transition probabilities across varying
levels of age and procrastination. These estimates illustrate how the
interaction between age and procrastination influences the likelihood of
progressing between cognitive states over time. Notably, while transition
probabilities remain relatively stable at very low levels of procrastination,
substantial shifts emerge as both age and procrastination increase. In
particular, older individuals with higher procrastination scores show an
elevated probability of cognitive decline transitioning from normative cognitive
function to mild cognitive impairment (MCI) and a reduced likelihood of
transitioning back from MCI to normative cognitive function, highlighting the
compounded risk posed by these two variables in later life.

```{r}
#| label: fig-age-preds
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Predicted transition probabilities by procrastination and age. *Note.* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset. Specific ages of interest (70, 80, and 90) are highlighted."
#| include: true

# Read in plot
fig_4 <- readRDS(here::here("files/figure_4.RDS"))

# Display plot
fig_4
```

# Discussion

Dementia poses a growing global health burden, with modifiable risk factors and
prodromal markers offering important targets for intervention [@livingston2024;
@livingston2020]. While apathy has been well established as a prodromal marker
and predictor of dementia [@donovan2014; @fresnais2023; @salem2023;
@vandalen2018], emerging evidence suggests that procrastination may also signal
early cognitive dysfunction [@friden2020; @zhang2019]. This study examined
whether procrastination, particularly in later life, could serve as a predictor
of cognitive decline. We assessed both cross-sectional differences in
procrastination across three cognitive groups (normative cognitive function, MCI
and dementia) and longitudinal associations with transitions in cognitive
status, to clarify whether procrastination functions as a novel and
age-sensitive marker of emerging cognitive impairment.

Our analysis revealed significant group differences in procrastination, with
individuals experiencing cognitive impairment, both MCI and dementia, reporting
higher procrastination scores than those with normative cognitive function.
These findings support the hypothesis that procrastination may be associated
with cognitive decline and aligns with emerging evidence linking procrastination
to cognitive dysfunction [@friden2020; @zhang2019]. Interestingly, while both
MCI and dementia groups exhibited elevated procrastination scores, no
significant difference emerged between these two groups. This suggests that
increases in procrastination may occur relatively early in the neurodegenerative
process, potentially preceding or paralleling the emergence of more overt
cognitive symptoms.

However, it should be noted that the number of participants classified as having
MCI $(n = 67)$ or dementia $(n = 27)$ was relatively small compared to those
with normative cognitive function $(n = 455)$. This imbalance in group sizes may
have limited the sensitivity of our comparisons and raises the possibility that
subtle differences between MCI and dementia groups were obscured by statistical
power constraints. Future research should aim to replicate these findings in
more evenly distributed samples to better determine whether procrastination
continues to increase across progressive stages of impairment or plateaus
following early decline.

Beyond these cross-sectional differences, our discrete-time Markov analysis
offered insight into how procrastination alongside other demographic and
psychological factors, influences the likelihood of transitioning between
cognitive states over time. Notably, procrastination emerged as a dynamic
predictor of cognitive change, interacting with age to significantly increase
the odds of transitioning from normative cognitive function to MCI, while
simultaneously decreasing the odds of moving from MCI back to normative
cognitive function.

These findings imply that procrastination may serve dual roles, both as a marker
of early cognitive dysfunction and, potentially, as a behavioral impediment to
moving from MCI state to the normative cognitive state. One plausible pathway
involves the reinforcement of maladaptive behaviors such as reduced cognitive
engagement or reduced participation in cognitively protective activities, such
as physical exercise, social interaction or medical adherence [@hajek2025;
@kelly2021; @sirois2007; @stead2010]. In parallel, procrastination has been
associated with chronic stress and elevated cortisol levels [@sirois2023a],
which may accelerate hippocampal atrophy, β-amyloid plaque deposition, and brain
inflammation [@franks2021; @wallensten2023], further undermining cognitive
resilience. These behavioral and biological mechanisms align with
self-regulation theories suggesting that procrastination reflects executive
dysfunction [@rozental2014; @steel2007], an early feature of neurodegenerative
progression [@clark2012].

The observed interaction with age (see @fig-age-preds) suggests that the
influence of procrastination on cognitive trajectories may grow stronger with
advancing age, a period during which neuro-plasticity diminishes, and behavioral
risk factors exert greater influence [@livingston2020; @livingston2024]. This
effect was particularly pronounced among the oldest participants. As illustrated
in @fig-age-preds, the effect of procrastination is relatively modest for
younger-older adults (those age 70). However, the slope of these transitions
steepens considerably for individuals aged 80 and especially for those age 90.
These findings indicate that procrastination is increasingly associated with a
higher risk of cognitive decline and a lower likelihood of improvement in
cognitive status in late old age.

This pattern underscores the possibility that procrastination functions as a
compounding risk factor in older adulthood, particularly among the oldest
individuals, by both reflecting emerging cognitive difficulties and potentially
accelerating their progression. In younger-old adults, cognitive reserve and
compensatory mechanisms [@gooijers2024] may buffer against the impact of poor
self-regulation. However, as individuals reach more advanced ages, these
protective systems weaken [@roberts2015]. Consequently, maladaptive behavioral
tendencies like procrastination can exert a disproportionate toll on cognitive
function. These findings underscore the potential value of addressing behavioral
regulation and self-management in older adults as part of broader dementia
risk-reduction strategies, particularly for those of a much older age.

By revealing a significant interaction between age and procrastination, this
study highlights the importance of broadening dementia risk models to
incorporate dynamic lifestyle and psychological variables. Procrastination, as
both a modifiable and measurable behavioral tendency [@vaneerde2018], represents
a promising target for low-cost, non-invasive interventions aimed at enhancing
cognitive resilience in older populations. Tracking self-regulatory behavioral
tendencies such as procrastination may offer an early warning system for
emerging cognitive risk, opening avenues for preventative action before
irreversible decline takes hold.

## Limitations

Despite the insights offered by this study, several limitations should be
considered. Most notably, although our longitudinal Markov modelling approach
captured changes in cognitive status, a key constraint was the measurement of
procrastination at only a single time point (2020). As a result, we assumed
temporal stability in procrastination scores across time and used this measure
as both a retrospective and prospective proxy for procrastination scores.
However, this precluded analysis of within-person changes in this behavior over
time and whether such changes affect cognitive transitions. It should be noted
that this limitation is inherent to the HRS dataset rather than a methodological
oversight (Juster & Suzman, 1995). Future studies should prioritise repeated
assessments of procrastination to determine whether increases in procrastination
precedes, accompanies, or follows cognitive decline. Moreover, our use of a
proxy measure of apathy may not have fully captured the multidimensional nature
of apathy. Future research should incorporate more validated apathy assessments
to strengthen the behavioral inferences drawn.

Additionally, as noted earlier, the relatively small number of individuals in
the MCI and dementia groups may have constrained statistical power for certain
comparisons and increased the potential for misclassification bias. Finally,
while the models adjusted for several key demographic and psychological
covariates, unmeasured confounding (undiagnosed medical conditions, medication
use, or sleep quality) could also have influenced cognitive outcomes. Future
research should seek to replicate and extend these findings using more
clinically diverse samples.

## Conclusions

In summary, this study offers preliminary evidence that procrastination may
function as an early behavioral tendency marker of cognitive decline,
particularly in older age. Individuals with MCI and dementia reported higher
procrastination scores, and longitudinal modelling revealed that
procrastination, especially when coupled with advancing age, was predictive of
increased cognitive decline. These findings underscore the importance of
everyday self-regulatory behaviors in dementia risk and resilience. As a
modifiable and measurable construct, procrastination holds promise as a target
for early detection and preventative intervention strategies aimed at sustaining
cognitive health in ageing populations.

\newpage

::: {.content-visible when-format="html"}
# References
:::
